Pairwise-complete correlation considered dangerous

### 2015-05-26 missing.Rmd
missing

The `use="complete"` option is one way to deal with missing values. It simply removes rows of the matrix `x` with missing observations. Since the third through fifth rows of our example matrix
are all identical we expect perfect correlation across the board, and indeed:
```{r}
cor(x, use="complete")
```
Reasonable people might question this approach. Deleting two observations has a huge effect on the correlation between `x[,1]` and `x[,2]` in this example because those columns exhibit large variance. The result really says that we should collect more observations!

@@

If you really don't want to do that, consider imputing the missing values. The simplest available method replaces missing values in each column with the mean of the non-missing values in the respective column:
```{r}
m = mean(na.omit(x[,3]))
xi = x
xi[is.na(x)] = m
cor(xi)
```
This can be done really efficiently when "centering" a matrix by simply replacing missing values of the centered matrix with zero.

@@

A more expensive approach bootstraps the missing values from the non-missing ones, either per column or from all the values if you expect them to be similarly distributed. You can decide by examining the empirical distributions of the non-missing data in each column.
```{r}
i = is.na(x[,3])
N = sum(i)
b = replicate(500, {x[i,3] = sample(x[!i,3], size=N, replace=TRUE);cor(x[,1:2],x[,3])})

@@

If you have lots of observations, consider partitioning them with a basic clustering algorithm first and then imputing the missing values from their respective cluster cohorts.

>>>>

The `use="complete"` option is one way to deal with missing values. It simply removes rows of the matrix `x` with missing observations. Since the third through fifth rows of our example matrix
are all identical we expect perfect correlation across the board, and indeed:
```{r}
cor(x, use="complete")
```
Reasonable people might question this approach. Deleting two observations has a huge effect on the correlation between `x[,1]` and `x[,2]` in this example mostly because of the large change in `x[,1]`. The result really says that we should collect more observations!

@@

If you really don't want to do that, consider imputing the missing values. The simplest method replaces missing values in each column with the mean of the non-missing values in the respective column:
```{r}
m = mean(na.omit(x[,3]))
xi = x
xi[is.na(x)] = m
cor(xi)
```
This can be done really efficiently when "centering" a matrix by simply replacing missing values of the centered matrix with zero.

@@

Another basic approach bootstraps the missing values from the non-missing ones:
```{r}
i = is.na(x[,3])
N = sum(i)
b = replicate(500, {x[i,3] = sample(x[!i,3], size=N, replace=TRUE);cor(x[,1:2],x[,3])})

@@

If you have lots of observations, consider partitioning them with a basic clustering algorithm first and then imputing the missing values from their respective cluster cohorts. Or consider a matching method or a regression-based imputation method. See the references below for many more details.
